---
title: "leopard"
author: "Charles T T Edwards (NIWA, Wellington, New Zealand)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{leopard-properties}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The `leopard` package - emergent properties

This document outlines the emergent properties of the `leopard` package. Because the package implements a stochastic model, these are best illustrated through simulation.

```{r}
library(leopard)
library(reshape2)
library(ggplot2)
```

## Initialisation

First define the basic demographic information and dimensions.
```{r}
# survival rates per demographic category
#ss <- c(0.3270833, 0.7197452, 0.9615385, 0.8882435, 0.9729730, 0.9382353, 0.9230769, 0.7219583, 0.9124424, 0.9642857, 1.0000000, 1.0000000, 0.9000000, 0.2857143)
ss <- c(0.4610291, 0.7197452, 0.9615385, 0.8882435, 0.9729730, 0.9382353, 0.9230769, 0.7219583, 0.9124424, 0.9642857, 1.0000000, 1.0000000, 0.9000000, 0.2857143)

# litter size per maternal age class
ll <- c(2, 2, 2, 2, 2)

# initial population numbers
x.initial <- c(nc  = 406,
               nj  = 201,
               saf = 92,
               f36 = 67,
               f48 = 63,
               f60 = 62,
               f72 = 93,
               f84 = 469,
               sam = 39,
               m36 = 62,
               m48 = 47,
               m60 = 78,
               m72 = 84,
               m84 = 246)

# number of iterations for simulation
niter <- 1e2
```

## Harem size

We can examine the harem size. 
```{r}
xx <- leopard(x.initial, survival.rates = ss, litter.sizes = ll)
harem(xx)@harem.size
```

## Survival

We can examine the distribution of realised cub survival rates when there is no harvest, which should match the empirically observed value of 0.3.
```{r}
xx <- leopard(x.initial, survival.rates = ss, litter.sizes = ll)
```

```{r, fig.width=6}
simulated.survival <- numeric(1e4)
for (i in 1:length(simulated.survival)) 
    simulated.survival[i] <- infanticide(survival(xx))@realised.survival.rate[1]

hist(simulated.survival, prob = TRUE)
abline(v = 0.3270833, col = 2)
```

## Infanticide

Cub survival is dependent on the number of adult males and the adult male survival. If there are a large number of males that die frequently, then infanticide rates will be high and cub survival will be low. Conversely, if there are either few males or they have high survivorship, infanticide rates will be low. 

Through simulation, we examine the relationship between male survival (adjusted via a constant harvest rate) and the probability of infanticide, and male survival and the cub survivorship.
```{r, fig.width=6, warning=FALSE}
harvest.rate <- seq(0, 0.4, length = 11)
selectivity  <- c(0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1)

prob.inf <- array(0, dim = c(length(harvest.rate), niter, 5))
cub.surv <- array(0, dim = c(length(harvest.rate), niter, 5))

# create new leopard object
xx <- leopard(x.initial, survival.rates = ss, litter.sizes = ll)
    
# loop over harvest rates
for (i in 1:length(harvest.rate)) {
    
    # loop over iterations
    for (j in 1:niter) {
        
        # re-initialise
        yy <- xx
        
        yy <- survival(yy)
        yy <- infanticide(yy)
        yy <- transition(yy)
        
        # project forward to adjust numbers
        for (k in 1:50) {
            
            yy <- birth(yy)
            
            zz <- harvest(yy, list(list(rate = harvest.rate[i], preference = selectivity)))
            
            yy <- survival(yy, zz[[1]]@kills)
            yy <- infanticide(yy)
            
            # for every 10th year
            if (k %% 10 == 0) {
                # record probability of infanticide
                # and cub survival
                prob.inf[i, j, k/10] <- yy@prob.infanticide
                cub.surv[i, j, k/10] <- yy@expected.survival.rate[1] * (1 - yy@prob.infanticide)
            }
                
            yy <- transition(yy)
            
        }
    }
}

# plot median probability of infanticide
prob.inf <- apply(prob.inf, c(1, 3), median)
dimnames(prob.inf) <- list(H = harvest.rate, Year = seq(10, 50, length = 5))
ggplot(melt(prob.inf)) + geom_line(aes(H, value, col = as.factor(Year))) + 
    ggtitle('Probability of infanticide') + labs(y = '', col = 'Year')

# plot median cub survivorship
cub.surv <- apply(cub.surv, c(1, 3), median)
dimnames(cub.surv) <- list(H = harvest.rate, Year = seq(10, 50, length = 5))
ggplot(melt(cub.surv)) + geom_line(aes(H, value, col = as.factor(Year))) + 
    ggtitle('Expected cub survival') + labs(y = '', col = 'Year')
```

## Hunting

We can examine the distribution of waiting times, and the realised hunting preference to ensure things are behaving as expected. 
```{r}
xx <- leopard(x.initial, survival.rates = ss, litter.sizes = ll)
```

```{r, fig.width=6}
simulated.days <- numeric(niter)
for (i in 1:length(simulated.days)) 
    simulated.days[i] <- implementation(xx, list(list(size = 1)))[[1]]@days[]

hist(simulated.days, prob = TRUE)
```

```{r, fig.width=6}
simulated.age.class <- numeric(niter)
for (i in 1:length(simulated.age.class)) 
    simulated.age.class[i] <- which(implementation(xx, list(list(size = 1, preference = "Braczkowski")))[[1]]@kills[] > 0)

hist(simulated.age.class, prob = TRUE)
```




